Expert-In-The-Loop Causal Discovery: Iterative Model Refinement Using Expert Knowledge

### 2024-11-26 paper/main.tex

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and informs effective interventions or policy decisions.
\emph{Causal Discovery} methods aim to uncover these relationships among random
variables from observational data. Methods for causal discovery have been
developed in both the Directed Acyclic Graphs (DAGs) and Structural Equation
Models (SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, numerous algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}.

In contrast, the SEM literature has focused more on developing tools to assist
researchers in manually constructing or modifying models while allowing them to
integrate their expert knowledge into the model-building process. Researchers
typically develop an initial model based on their expertise and then use
various tools to guide modifications to improve the model's fit to data. This
model construction process is commonly known as \emph{Specification Search}
\citep{Long1983} and uses method like modification indices, and Wald-based
tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood. Moreover, their output can vary
		significantly depending on the algorithm and their
		hyperparameters. Combined with the lack of robust methods to
		test these algorithms' performance on a given dataset, it can
		be difficult to trust their outputs. Consequently, researchers
		usually have to rely on post-hoc modifications to the model
		based on their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Figure~\ref{fig:intro} shows some of these issues when applied to the Adult
Income dataset. Due to these challenges, whether constructing the model
manually or using an algorithm, we typically have to rely on some expert
knowledge to get a model that we can use for further analysis. Hence, it is
crucial to have tools to guide this construction or modification process.

@@

>>>>

Understanding cause-and-effect relationships between variables is a fundamental
goal in many scientific inquiries, as it reveals the underlying mechanisms of
observed phenomena and help in determining effective interventions or policy
decisions. \emph{Causal Discovery} methods aim to uncover these relationships
among random variables from observational data. Methods for causal discovery
have been developed in both the Directed Acyclic Graphs (DAGs) and Structural
Equation Models (SEMs) literatures, each adopting a different approach.

In the DAG literature, the focus has primarily been on developing automated
algorithms to learn causal structures from datasets. As a result of these
efforts, many algorithms for causal discovery have been proposed, such as
constraint-based methods like the PC algorithm \citep{Spirtes2001} and Fast
Causal Inference (FCI) \citep{Spirtes2000}), score-based methods like
Hill-Climb Search and Greedy Equivalence Search (GES) \citep{Chickering2002},
and continuous optimization-based methods like NO TEARS \citep{Zheng2018} and
DAGMA \citep{Bello2022}. In contrast, the SEM literature has focused more on
developing tools to assist researchers in manually constructing or modifying
models while allowing them to integrate their expert knowledge into the
model-building process. Researchers typically develop an initial model based on
their expertise and then use various tools to guide modifications to improve
the model's fit to data. This model construction process is commonly known as
\emph{Specification Search} \citep{Long1983} and uses method like modification
indices, and Wald-based tests \citep{Marcoulides2018}.

@@

\begin{enumerate}
	\item \textbf{Lack of Trust:} Although most algorithms are
		asymptotically consistent, their finite sample properties are
		not well understood. Moreover, their output can vary
		significantly depending on the chosen algorithm and their
		hyperparameters. Combined with the lack of robust methods to
		test these algorithms' performance on a given dataset, it can
		be difficult to trust their outputs. Consequently, researchers
		usually have to rely on post-hoc modifications to the model
		based on model testing and their domain expertise.
	\item \textbf{Outputs Markov Equivalence Class (MEC):} Multiple
		DAGs can be faithful to a given observational dataset. Thus,
		using only observational data, algorithms can only recover the
		MECs that can contain a combination of directed and undirected
		edges. However, most methods for downstream tasks, such as
		identification or causal effect estimation, assume knowledge of
		a fully oriented DAG. As a result, researchers have to manually
		orient these edges before using them for analysis.
\end{enumerate}

Figure~\ref{fig:intro} shows an example of application of a causal discovery
algorithms namely PC on the Adult Income dataset. \todo[inline]{Explain}.

Due to these challenges, whether constructing the model manually or using an
algorithm, we typically have to rely on some expert knowledge to get a model
that we can use for further analysis. Hence, it is crucial to have tools to
guide this construction or modification process.

@@

\subsection{Ranking using Correlation Coefficients}
